Gradual Soundness: Lessons from Static Python

### 2021-10-28 ecoop-2022/paper.tex

\begin{itemize}
  \item
    Evidence that gradual soundness delivers fast performance incrementally,
    as programmers migrate to sound types.
    \SP{} can be applied to a codebase with no refactoring and no performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description and analysis of the \SP{} core language.
    The analysis includes a proof of type soundness,
    a proof that optimization soundly improves code,
    and [FILL perhaps] a metric to quantify the refactoring cost and optimization benefit
    of a type migration.
  \item
    In-depth discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

No way to write the dynamic type.
Instead, it fills in for absent annotations.
Either totally missing or a missing generic parameter.

%% TODO are there any types C[T0, T1] where you might want C[dyn, T1]? how to express?

Cannot appear everywhere ... ChkDict can't hold a Dynamic,
and primitives are incompatible with Dynamic.

@@

\subsection{Benchmarks}

@@

@@

>>>>

\begin{itemize}
  \item
    Evidence that gradual soundness delivers fast performance incrementally,
    as programmers migrate to sound types.
    \SP{} can be applied to a codebase with no refactoring and no performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description of the \SP{} core language and an analysis of its soundness
    and optimizations.
    Soundness is not uniform across all types.
    Some are fully sound while others are constructor sound.
  \item
    In-depth discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

Can't really write the dynamic type.
But ANY unbound id. or non-type object gets interpreted as the dynamic type for now.

Cannot appear everywhere ... ChkDict can't hold a Dynamic (or rather, you can declare this type and cannot instantiate it),
and primitives are incompatible with Dynamic.

@@

\subsection{Microbenchmarks}

richards etc.

@@

\section{Design Principles}

\subsection{Nominal, Checked Types}

CheckedDict enables strong type checks and optimizations, but is incredibly painful to use.
It is painful even though ChkDict supports the full dict API
(any context that uses a Python dict can use a ChkDict without code changes).

The problem is that the ChkDict type rejects all Python values.
It accepts only ChkDict values, built through a special constructor
that installs a tag for type tests and guards writes.
Suppose that \code{f} is an untyped dict function.
As is, it can process ChkDict but gets no benefit from type checks nor from optimizations.
Adding a ChkDict domain type enables optimizations within \code{f} but raises a non-local
problem: all callers of \code{f} must be sure to create a ChkDict.
These ChkDicts must also be monomorphic to match whatever type \code{f} chose.
It can take many edits to get a program running again after adding a ChkDict type.

Lessons:
\begin{itemize}
  \item
    Nominal types are not compatible with structual values, such as Python values.
    Programmers are forced to edit old code to use such types ... unless the recent Nom work has a better idea~\cite{mt-oopsla-2021}
  \item
    Monomorphic nominal types are even worse.
    Editing old code to fit their rigid constraints may not be feasible.

@@

Use confined GT to relax ChkDict ... let programmers decide whether
the type should reject or convert untyped data.
Maybe a ChkDict function could compile to an "overloaded" version with
fast and slow paths for ChkDict vs normal dict.
